To Batch or Not to Batch: Test-Ordering Variability in the ED

### 2024-09-09 manuscript/manuscript.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variabilityin the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing. Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy. Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

@@

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe approximately $48,000$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include urinary complaints and allergic reactions. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving physicians who treat over 500 ED cases per year.

@@

  *(c) 72-hour return*-- The rate of patients returning to the ED within 72 hours of their initial visit is a proxy for the quality of care received. A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit. We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge. The hypothesis is that batch testing may lead to variations in the 72-hour return rate, potentially influencing the quality of care received.

@@

We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window. Sensitivity analyses on this cutoff point showed that our results are robust to this definition. This operational definition aligns with the concept of batching as a single comprehensive effort. For our analysis, all lab-based testing ordered in this timeframe are categorized as one distinct type of testing (lab). In contrast, each imaging test (e.g., X-ray, CT scan) are considered a separate distinct test. Therefore, a batch in our study is defined as consisting of two or more distinct diagnostic tests, which could be a combination of lab and imaging tests or multiple imaging tests.

@@

\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period. On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load. Key physiological markers such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients, albeit at varying degrees, highlighting critical aspects of emergency care. Notably, a small proportion of patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the diversity of cases encountered in the ED setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting that most cases fall within the moderate acuity level.

@@

\begin{table}[ht]
\centering
\caption{Summary Statistics}
\label{tab:summary_statistics}
\begin{tabular}{p{10.5cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
ED Volume & 24.18 & 15 & 25 & 32 \\
Tachycardic & 0.19 & & & \\
Tachypneic & 0.09 & & & \\
Febrile & 0.02 & & & \\
Hypotensive & 0.01 & & & \\ 
ESI & 2.78 & 2 & 3 & 3 \\
Complaint: Abdominal & 0.14 & & & \\
Complaint: Extremity & 0.12 & & & \\
Complaint: Chest Pain & 0.08 & & & \\
Complaint: Neurological & 0.08 & & & \\
Complaint: Gastrointestinal & 0.08 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.33 & 44 & 62 & 75 \\
Race: White & 0.89 & & & \\
Race: Black & 0.04 & & & \\
Race: Asian & 0.03 & & & \\
Gender: Female & 0.54 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.47 & & & \\
Ultrasound & 0.11 & & & \\
Non-Contrast CT & 0.21 & & & \\
Contrast CT & 0.19 & & & \\
Lab & 0.77 & & & \\
Tests were Batch Ordered & 0.38 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{table}

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
\end{equation}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically. \hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $8$pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 50 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Physician for High Volume Complaints}
  \label{fig:physician_batching}
  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
\begin{tablenotes}
\tiny
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests. Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue). The quartile of batchers is calculated based on the batching rate across all complaint areas per physician. Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching. Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
\end{tablenotes}  
\end{figure}

\begin{table}[ht]
\centering
\caption{First Stage: Effect of Batch Tendency on Test Batching}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43327 & 43327 & 43327 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

\begin{table}[htbp]
\centering
\caption{Placebo Exercise: Reduced-Form Results for Rarely Batched Samples}
\label{tab:placebo_check}
\begin{tabular}{p{8cm}ccc}
\toprule
& ln(LOS) & Number of Tests & 72 Hour Return \\ 
& (1) & (2) & (3) \\
\midrule
Batch Tendency & $-0.465$ & $-0.699$ & $-0.011$ \\
& $(0.645)$ & $(0.427)$ & $(0.095)$ \\
                    &     &     &  \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Testing inclination? & Yes & Yes & Yes \\
                    &     &     &  \\
N & 2,361 & 2,361 & 2,361 \\
\addlinespace
\midrule
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field. Estimates come from regression on conditions that are batched less than 10 percent of the time. See text for residualization fixed effects and baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

In this section we examine the causal effects of batch ordering in the ED. Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS, number of tests, and 72-hour return.

\begin{table}[!htbp] \centering 
  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
  \label{tab:2SLS} 
\begin{tabular}{p{7cm}ccc}
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & ln(LOS) & Number of Tests & 72 Hour Return \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Batch & $-$0.086 & 0.135$^{***}$ & 0.018 \\ 
  & (0.419) & (0.046) & (0.018) \\ 
  & & & \\ 
 Testing Inclination & 0.210$^{**}$ & 0.335$^{***}$ & $-$0.009$^{***}$ \\ 
  & (0.082) & (0.008) & (0.003) \\ 
  & & & \\ 
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
\textit{N} & 41,929 & 41,929 & 41,929 \\ 
R$^{2}$ & 0.199 & 0.312 & 0.008 \\ 
Adjusted R$^{2}$ & 0.194 & 0.307 & 0.002 \\ 
Residual Std. Error (df = 41667) & 0.496 & 0.817 & 0.184 \\ 
\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the IV model are for the baseline sample described in the text. Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

*ED length of stay (LOS).*-- Our analysis reveals that batch ordering does not significantly reduce the length of stay in the ED. This finding holds even after accounting for various control variables, including ED volume and patient acuity. Contrary to common assumptions that batching tests could expedite patient throughput, our results suggest that the efficiency gains could be offset. A possible explanation lies in excess tests being ordered due to batching, which could lead to increased wait times for patients. This is consistent with our finding that batching increases the number of tests ordered.

*Resource Utilization.*-- The impact of batching on resource utilization is pronounced. We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 2,206 additional tests during our study period. This is likely due to the information gain that occurs from sequential test ordering, where a physician may order additional tests after receiving the results of the first test.

*72-hour Return*-- Our findings indicate that batch ordering does not significantly impact the likelihood of a 72-hour return. This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates. This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.

The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED. While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization without corresponding gains in patient throughput or reductions in short-term return rates. These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.

@@

Our results highlight that changes in resource utilization and costs can arise as a consequence of small variations in physician test ordering strategies in a single medical encounter, at the emergency department. Our instrumental variable approach suggests that the causal effects of batching are substantial: 13.5 excess tests arise per 100 batched orders. This is particularly pronounced in high-volume complaint areas, where batching leads to a significant increase in the number of tests ordered. However, we find no significant impact of batching on ED length of stay or 72-hour return rates. This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates. This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.

@@

\begin{table}[!htbp]
\centering 
  \caption{Testing the Monotonicity Assumption} 
\label{tab:monotonicity}
\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
\hline \\[-1.8ex] 
Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\tiny
\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample. Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

>>>>

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing. Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy. Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

@@

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe approximately $48,000$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include urinary complaints and allergic reactions. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving full-time physicians who treat over 500 ED cases per year.

@@

@@

We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window. In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition. This operational definition aligns with the concept of batching as a single comprehensive effort. For our analysis, all lab-based testing ordered in this timeframe are categorized as one distinct type of testing (lab). In contrast, each imaging test (e.g., X-ray, CT scan) are considered a separate distinct test. Therefore, a batch in our study is defined as consisting of two or more distinct diagnostic tests, which could be a combination of lab and imaging tests or multiple imaging tests.

@@

\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period. Data is at the patient encounter level. On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load. Key physiological markers such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients, albeit at varying degrees, highlighting critical aspects of emergency care. Notably, a small proportion of patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the diversity of cases encountered in the ED setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting that most cases fall within the moderate acuity level.

@@

\begin{table}[ht]
\centering
\caption{Summary Statistics of Emergency Department Encounters}
\label{tab:summary_statistics}
\begin{tabular}{p{10.5cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
ED Volume & 24.18 & 15 & 25 & 32 \\
Tachycardic & 0.19 & & & \\
Tachypneic & 0.09 & & & \\
Febrile & 0.02 & & & \\
Hypotensive & 0.01 & & & \\ 
ESI & 2.78 & 2 & 3 & 3 \\
Complaint: Abdominal & 0.14 & & & \\
Complaint: Extremity & 0.12 & & & \\
Complaint: Chest Pain & 0.08 & & & \\
Complaint: Neurological & 0.08 & & & \\
Complaint: Gastrointestinal & 0.08 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.33 & 44 & 62 & 75 \\
Race: White & 0.89 & & & \\
Race: Black & 0.04 & & & \\
Race: Asian & 0.03 & & & \\
Gender: Female & 0.54 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.47 & & & \\
Ultrasound & 0.11 & & & \\
Non-Contrast CT & 0.21 & & & \\
Contrast CT & 0.19 & & & \\
Lab & 0.77 & & & \\
Tests were Batch Ordered & 0.38 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{table}

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \varepsilon_{i,t}
\end{equation}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically. \hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $6$pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 94 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Physician for High Volume Complaints}
  \label{fig:physician_batching}
  \includegraphics[width=1\textwidth]{../outputs/figures/batch_ordering.png}
\begin{tablenotes}
\small
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests. Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue). The quartile of batchers is calculated based on the batching rate across all complaint areas per physician. Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching. Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
\end{tablenotes}  
\end{figure}

\begin{table}[ht]
\centering
\caption{First Stage: Effect of Batch Tendency on Test Batching}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 1.06 *** & 1.06 *** & 1.06 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 23.87 & 94.41 & 90.44  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 41929 & 41929 & 41929 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

\begin{table}[htbp]
\centering
\caption{Placebo Exercise: Reduced-Form Results for Rarely Batched Samples}
\label{tab:placebo_check}
\begin{tabular}{p{8cm}ccc}
\toprule
& ln(LOS) & Number of Tests & 72 Hour Return \\ 
& (1) & (2) & (3) \\
\midrule
Batch Tendency & $-0.465$ & $-0.699$ & $-0.011$ \\
& $(0.645)$ & $(0.427)$ & $(0.095)$ \\
                    &     &     &  \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Testing inclination? & Yes & Yes & Yes \\
                    &     &     &  \\
N & 2,361 & 2,361 & 2,361 \\
\addlinespace
\midrule
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field. Estimates come from regression on conditions that are batched less than 10 percent of the time. See text for residualization fixed effects and baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

In this section we examine the causal effects of batch ordering in the ED. Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS and number of tests ordered. We first start by estimating the effects of batch ordering on the number of tests ordered using the standard 2SLS approach. Next we estimate the direct and indirect effects of batch ordering on LOS using a causal mediation analysis.

\begin{table}[!htbp]
\centering
\begin{threeparttable}
\caption{Combined 2SLS and Causal Mediation Analysis Results}
\label{tab:combined_results}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
\toprule
& \multicolumn{2}{c}{2SLS Results} \\
\cmidrule{2-3}
& ln(LOS) & Number of Tests \\
\midrule
Batch & -0.086 & 0.135$^{***}$ \\
& (0.419) & (0.046) \\
& & \\
\addlinespace
Seasonality and Shift Fixed Effects & Yes & Yes \\
Chief Complaint and Severity & Yes & Yes \\
\midrule
Observations & 41,901 & 41,901 \\
R$^{2}$ & 0.183 & 0.310 \\
\bottomrule
\end{tabular*}

\smallskip % Adds some space before the next part of the table

\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
\toprule
& \multicolumn{2}{c}{Causal Mediation Analysis} \\
\cmidrule{2-3}
Effect & Estimate & 95\% CI \\
\midrule
ACME & 0.036 & [-0.009, 0.08] \\
ADE & -0.136 & [-0.221, -0.04] \\
Total Effect & -0.100 & [-0.195, 0.00] \\
Proportion Mediated & -0.333 & [-2.831, 1.21] \\
\bottomrule
\end{tabular*}

\begin{tablenotes}
\small
\item The 2SLS results are based on robust standard errors clustered at the physician level, with controls for seasonality, shift fixed effects, chief complaint, and severity. The causal mediation analysis examines whether the number of tests mediates the effect of batching on length of stay, using a quasi-Bayesian confidence interval approach with 1000 simulations.
\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
\end{tablenotes}
\end{threeparttable}
\end{table}

*ED length of stay (LOS).*-- Our causal mediation analysis reveals that the Average Direct Effect (ADE) of test batching on the natural log of LOS is -0.13607, indicating that, when not considering the mediating effect of the number of tests, batching leads to a decrease in the expected LOS. Specifically, this translates to an approximate 12.7 percent reduction in LOS when tests are batched, all else being equal. This result, complemented by our 2SLS estimate of -0.086, suggests that the direct effect of batch ordering on LOS is negative. Our results also suggest a "washing out" effect, where the effeciency gains from batching are offset by the increased number of tests ordered. This is consistent with our finding that batching increases the number of tests ordered.

*Resource Utilization.*-- The impact of batching on resource utilization is pronounced. We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 13.5 excess tests per 100 batch orders, where excess means tests that would not have been ordered had the physician not batched. This result is consistent with the notion that batching leads to increased resource utilization in the ED.

The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED. While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out. These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.

@@

Our results highlight that changes in resource utilization and costs can arise as a consequence of small variations in physician test ordering strategies in a single medical encounter, at the emergency department. Our instrumental variable approach suggests that the causal effects of batching are substantial: 13.5 excess tests arise per 100 batched orders. This is particularly pronounced in high-volume complaint areas, where batching leads to a significant increase in the number of tests ordered. However, we find no significant impact of batching on ED length of stay. Causal mediation analysis reveals that the direct effect of batching on length of stay is negative, but this effect is offset by the increased number of tests ordered. This suggests that the efficiency gains from batching are washed out by the increased resource utilization.

Our findings have important implications for emergency care practice. While batching is often seen as a method to enhance efficiency, our results suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out. These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing. Future research should explore the impact of batching on other outcomes, such as patient satisfaction and clinical outcomes, to provide a more comprehensive understanding of the implications of batching in the emergency department.

@@

\begin{table}[!htbp]
\centering 
  \caption{Testing the Monotonicity Assumption} 
\label{tab:monotonicity}
\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
\hline \\[-1.8ex] 
Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample. Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}
